---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r} 
#Reading data
expeditions_data <- readr::read_csv('https://github.com/rfordatascience/tidytuesday/blob/master/data/2020/2020-09-22/expeditions.csv?raw=true')
library(forcats)
library(dplyr)

#15 most frequent peaks
freq_exped <- expeditions_data %>%
    mutate(common_peaks = fct_lump(peak_name, n = 15))

#Common peaks without other
final_exped <- subset(freq_exped, common_peaks!="Other")

#I used reorder. I couldn't use fct_ordert használni, it said that hte 'Number of expeditions' variable does not exists.
ordered_exped <- within(final_exped, 
                   common_peaks <- factor(common_peaks, 
                                      levels=names(sort(table(common_peaks), 
                                                        decreasing=FALSE))))

#ggplot 
library(tidyverse)
  ggplot(ordered_exped, aes(y = common_peaks, fill = season)) + 
  geom_bar() + 
  scale_fill_viridis_d() +
  theme_set(theme_light()) +
  labs (y = NULL,
        x = "Number of expeditions",
        title = "The 15 most popular peaks stacked by season of expedition") +
  theme(legend.position="bottom")
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
phd_raw <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-19/phd_by_field.csv')
library(dplyr)
library(scales)
library(RColorBrewer)

# I removed NA
phd_tidy <- drop_na(phd_raw)

#aggregate data
all_phd <- phd_raw %>% group_by(broad_field) %>% count(year, wt = n_phds)

#ggplot
all_phd %>% ggplot() +
    aes(x = year, color = broad_field, y=n) + 
    geom_line(size = 1.2, show.legend = TRUE) +
    scale_x_continuous(name = NULL, 
                       breaks = scales::pretty_breaks(n= 4)) +
    scale_y_continuous(name = NULL,
                       labels = scales::comma_format(decimal.mark = ","))+
    labs( title = 'Number of awarded Ph.D.-s in the US by year',
          color = 'Broad field')+
    theme_minimal() + 
    scale_color_brewer(palette ="Dark2")

```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
bike_raw <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-11-05/commute.csv')
library(dplyr)
library(scales)
library(forcats)

#check data
table(bike_raw$state) #javítani kell a Ca-t Californiára
bike_raw$state[bike_raw$state=="Ca"]<-"California"

#aggregate data
tidy_bike <- bike_raw %>% group_by(state) %>% count(mode, state_region, state_abb, wt = n)
plot_data <- tidy_bike %>% pivot_wider(names_from = mode, values_from = n)

#ggplot
plot_data %>% ggplot(aes(x=Walk, y=Bike, color = state_region, label=state_abb)) + 
    geom_point(size = 2) +
    theme_light() + 
    scale_y_continuous(name = 'Number of ppl walking to work (log N)', trans = "log10", labels = scales::comma_format(decimal.mark = ",")) + 
    scale_x_continuous(name = 'Number of ppl biking to work (log N)', trans = "log10", labels = scales::comma_format(decimal.mark = ",")) +
    labs( title = 'Title number of people walking vs. biking to work in each USA state', color = 'State region') + 
  geom_text(aes(label=state_abb), color = "black")

#On the original picture not every point has a name, but I couldn't figure out what was the rule. 
```
