---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
install.packages("viridis")
install.packages("RColorBrewer")

```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
library(forcats)
library(dplyr)
library(ggplot2)
tuesdata <- tidytuesdayR::tt_load('2020-09-22')
expeditions <- tuesdata$expeditions
```


```{r}
library(viridis)

df <- expeditions %>% 
  mutate(peak_name = fct_lump(peak_name, n = 15)) %>%
  filter(peak_name != 'Other') %>%
  group_by(peak_name, season) %>%
  summarise(counts = n()) %>%
  ungroup() %>%
  arrange(desc(counts))

ggplot(df, aes(fct_reorder(peak_name, counts,.fun = sum), y = counts, fill = season)) + 
  geom_bar(stat = "identity") + 
  coord_flip() + 
  labs(title="The 15 most popular peaks stacked by season of expedition", x ="", y = "Number of expeditions") +    
  scale_color_viridis_d() + 
  scale_fill_viridis_d() +
  theme_light() 
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
phd_by_field <- tidytuesdayR::tt_load('2019-02-19')$phd_by_field
```

```{r}
library(RColorBrewer)

phds_by_broad_fields <- phd_by_field %>% 
  group_by(broad_field,year) %>% 
  summarise(n_phds = sum(n_phds, na.rm = TRUE))

ggplot(phds_by_broad_fields, aes(x=year, y=n_phds, group=broad_field, color=broad_field)) +  
  geom_line(size=1.2) + 
  scale_x_continuous(breaks=pretty_breaks(5)) +
  scale_y_continuous(labels = comma_format()) +
  labs(title="Number of awarded Ph.D.-s in the US by year", x ="", y = "") +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal() 
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
commute <- tidytuesdayR::tt_load('2019-11-05')$commute
```


```{r}
library(scales)
commute_by_state <- commute %>% 
  group_by(state_abb,state_region) %>% 
  summarise(bike = sum(n[mode == 'Bike'], na.rm = TRUE),walk = sum(n[mode == 'Walk'], na.rm = TRUE))

ggplot(commute_by_state, aes(x=walk, y=bike,group=state_region, color=state_region)) +
  geom_point(size=2) + 
    geom_text(aes(label=state_abb),color="black") +
  labs(title="Title number of people walking vs. biking to work in each USA state", x ="Number of ppl walking to work(log N)", y = "Number of ppl biking to work (log N)", color="State region") +
  scale_x_log10(breaks = log_breaks(n = 5),labels = comma_format()) +
  scale_y_log10(limits=c(25, 150000), breaks=10^(2:5),labels = comma_format())+
  theme_light() 
```
