---
title: 'Assignment 2: Data visualization'
author: "Tamas Nagy"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
library(ggplot2)
library(viridis)
library(dplyr, warn.conflicts = FALSE)
options(dplyr.summarise.inform = FALSE)
library(scales)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
#Assigning the data to a variable
expeditions <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2020/2020-09-22/expeditions.csv', col_types = cols())

#Getting the 15 most frequent peaks
top_peak <- fct_lump_n(f = expeditions$peak_name, n = 15)
exp_data <- subset(expeditions, peak_name == top_peak) #select variables

#Plotting data
ggplot(data = exp_data, 
  aes(y = fct_rev(fct_infreq(peak_name)), fill = season)) + 
  geom_bar() + 
  labs(x = "Number of expeditions", y = NULL, title = "The 15 most popular peaks stacked by season of expedition") +
  scale_fill_viridis(discrete = TRUE) +
  theme_light() +
  theme(legend.position = "bottom")
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

```{r}
#Assigning the data to a variable
phd_field <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-02-19/phd_by_field.csv", col_types = cols())

#Grouping and plotting
all_phd <- phd_field %>%
	group_by(broad_field, year) %>%
	summarize(n = sum(n_phds, na.rm = TRUE)) %>% 
  ggplot(aes(x = year, y = n, color = broad_field)) +
  geom_line(size = 1.2) +
  scale_x_continuous(breaks = scales::pretty_breaks(5)) + 
  scale_y_continuous(labels = comma_format()) +
  labs(x = NULL, y = NULL, title = "Number of awarded Ph.D.-s in the US by year", colour = "Broad field") +
  scale_color_brewer(palette = "Dark2") +
  theme_minimal()

#Calling all_phd
all_phd
```

## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

```{r}
#Assigning the data to a variable
commute_mode <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-11-05/commute.csv", col_types = cols())

#Aggregating all commutes
commute_agg <- commute_mode %>%
	group_by(state_region, state_abb, mode) %>%
  summarise(n = sum(n, na.rm = TRUE))

#Aggregating bikers
biker_agg <- commute_agg %>%
  group_by(state_region, state_abb, n) %>%
  filter(mode == "Bike") %>%
  rename(bikers = n)

#Aggregating walkers
walker_agg <- commute_agg %>%
  group_by(state_region, state_abb, n) %>%
  filter(mode == "Walk") %>%
  rename(walkers = n)

#Merging dataframes together 
commuters <- list(biker_agg, walker_agg) %>% 
  reduce(full_join, by = "state_abb") #reducing a list to a single value

#Plotting
ggplot(data = commuters, aes(x= walkers, y = bikers, color = state_region.x)) +
  geom_point(size = 2) +
  geom_text(aes(label=state_abb), colour = "black", size = 3, na.rm = TRUE) +
  scale_x_log10(labels = comma) +
  scale_y_log10(labels = comma) +
  labs(x = "Number of ppl walking to work (log N)", y = "Number of ppl biking to work (log N)", title = "Title number of people walking vs. biking to work in each USA state", color = "State region") +
  theme_light()
```
