---
title: 'Assignment 2: Data visualization'
author: "Adel Kovacs"
output: html_document
editor_options: 
  chunk_output_type: console
---

You will have to create 3 plots based on the datasets and instructions detailed below. You will find the plots themeselves in the `assignments/assignment_2_plots`. Your task is to write the code that will reproduce the plots as closely as possible.

# Skills needed to solve this assignment

-   Using R and RStudio, reading data
-   Reporting using RMarkdown
-   Using Git and Github (for submitting the task)
-   Data manipulation (e.g. dplyr, tidyr), and working with factors (forcats)
-   Data visuzlization (ggplot2)

*I have added some extra packages*

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(tidytuesdayR)
library(forcats)
library(ggplot2)
library(dplyr)
library(tidyr)
library(scales)
library(RColorBrewer)
```

## Task 1: Climbing expeditions

The 2020-09-22 TidyTueday datasets are about climbing expeditions. From the three datasets, use the "expeditions". Reproduce the plot below! Notice a few things:

-   Use `forcats::fct_lump()` to get the 15 most frequent peaks, and drop the "Other" category.
-   The bars are ordered by the sum of all expeditions (use `fct_reorder()`).
-   The bar colors use the viridis palette and light theme.

```{r}
tt_data <- tt_load("2020-09-22")
expeditions <- tt_data$expeditions
```

*I made sure the variables I need are handled as factors*
```{r}
expeditions$season <- as.factor(expeditions$season)
expeditions$peak_name <- as.factor(expeditions$peak_name)
```

*I have checked the peaks*

```{r}
fct_count(expeditions$peak_name)
```

*I choosed the 15 most popular and dropped the others*

```{r}
expeditions_to_plot <- expeditions %>%
  mutate(peak_name = fct_lump_n(peak_name, 15)) %>%
  filter (peak_name != "Other") %>%
  count (peak_name, season)
```

*Lastly I made the plot using the instructions*
*It is odd, because ever since now the order of the bars were correct. Although I have not changed anything in the code it is not the same as the example (or exact order) anymore. I would love to know why, so any feedback is appriciated*

```{r}
ggplot(expeditions_to_plot, aes(x = n, y = reorder(peak_name, n), fill = season)) +
  geom_col () +
  theme_light () +
  scale_fill_viridis_d() +
  labs (title = "The 15 most popular peaks stacked by season of expedition", x = "Number of expeditions", y = NULL) +
  theme (legend.position = "bottom")
```

## Task 2: PhDs awarded

The 2019-02-19 TidyTueday dataset is about phd-s awarded by year and field. There is only one dataset, it is called `phd_by_field`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all phd-s by broad fields.
-   To make the x axis breaks pretty, use `scales::pretty_breaks()`, to make the y axis labels comma formatted, use `scales::comma_format()`.
-   The line size is 1.2, the colors are from the brewer "Dark2" palette. The theme is set to minimal.

*First I have loaded the data and checked it*

```{r}
tt_data <- tt_load("2019-02-19")
phd_by_field <- tt_data$phd_by_field
phd_by_field
```

*Then I have changed some variable types to work with them*

```{r}
phd_by_field$broad_field <- as.factor(phd_by_field$broad_field)
phd_by_field$n_phds <- as.integer(phd_by_field$n_phds)
```

*After I made a tibble about the variables I want to plot*

```{r}
sum_by_group <- phd_by_field %>%
  filter(n_phds != "NA") %>%
  group_by(broad_field, year) %>%
  summarize (Count =sum(n_phds))
sum_by_group
```

*Lastly I made the plot*

```{r}
ggplot(sum_by_group, aes(x=year, y=Count, color=broad_field)) +
  geom_line (size=1.2) +
  scale_x_continuous(breaks = pretty_breaks(6)) +
  scale_y_continuous(label = comma_format()) +
  scale_colour_brewer(palette="Dark2")+
  theme_minimal() +
  labs(title = "Number of awarded Ph.D.-s in the US by year") 
```


## Task 3: Commute in the US

The 2019-11-05 TidyTueday dataset is about commuting to work in each city in the US by bike or on foot. There is only one dataset, it is called `commute`. Reproduce the plot below!

Notes:

-   First you have to aggregate the data to count all commutes by state.
-   Both axis scales are log transformed and the labels comma formatted, using `scales::comma_format()`
-   The point size is 2, . The theme is set to light.

*First I have loaded the data*

```{r}
tt_data <- tt_load("2019-11-05")
commute <- tt_data$commute
```

*Then I changed the type of the variables I want to work with*

```{r}
commute$mode <- as.factor(commute$mode)
commute$state_abb <- as.factor(commute$state_abb)
```

*After I made a dataframe with the variables and values I am interested in*

```{r}
commute_to_plot <- 
  commute %>%
  group_by(mode, state_abb, state_region) %>%
  summarize (Count =sum(n))
```

*To have all columns I need, I made it wider*

```{r}
  commute_to_plot_wider <- pivot_wider(commute_to_plot, names_from = "mode", values_from = "Count")
```

*Lastly I made the plot. Seems like I have a bit different data on, but I assume that maybe the data was updated since then*

```{r}
ggplot(commute_to_plot_wider, aes(x=Walk, y=Bike, color=state_region)) +
  geom_point (size= 2) +
  geom_text(aes(label=state_abb), colour = "black") +
  theme_light() +
  scale_x_log10 (label = comma_format())+
  scale_y_log10 (limits = c(100, 100000), label = comma_format()) +
  labs(title = "Title number of people walking vs. biking to work in each USA state") +
  xlab("Number of ppl walking to work (log N)") +
  ylab("Number of ppl biking to work (log N)") +
  scale_colour_discrete(name="State region")
```

