---
title: "Assignment 1 for R course"
author: "Kovacs_Adel"
date: "1/2/2022"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## PSZM21-MO-KUT-104:2	Komplex adatelemzési eljárások-Adatelemzés R-programnyelven

### The goal was to clean and tidy the databese and exploring it 

I have listed all the questions from the assignment, then answered them

First, I have loaded the packages I am going to need and came back here every time I needed to add something 
```{r, message = FALSE}
library(dplyr)
library(forcats)
library(tidyr)
library(readr)
library(splitstackshape) 
library(stringr)
library(rlang)
```

**1. Q: Read the data**

1. A: I loaded the dataset and checked it

```{r, message = FALSE}
cocktail_bar <- read_tsv(url("https://raw.githubusercontent.com/nthun/cocktail-balance/master/cocktail_data.tsv"))
View(cocktail_bar)
```

**2. Q: Transform the data table and clean the ingredient variable!**

2. A: I have realized the separator, so next I count how much is the most ingredients: 

```{r}
ingredients <- cocktail_bar$ingredients
str_count(ingredients, pattern = "<br/>")
```

5 is the most, so I am gonna need 6 columns for ingredients:

```{r, message = FALSE, warning = FALSE}
cocktail_bar_separated <- cocktail_bar %>%
  separate(ingredients,into= c("ingredient1", "ingredient2", "ingredient3", "ingredient4", "ingredient5", "ingredient6"), sep= "<br/>")
```

Cleaning the ingredient1:ingredient6 columns and checking the result

```{r}
cocktail_bar_cleaned <- cocktail_bar_separated %>% 
  mutate(across(ingredient1:ingredient6, ~sub(".*oz ", '', .))) %>%
  mutate(across(ingredient1:ingredient6, ~sub(".*OZ ", '', .))) %>% 
  mutate(across(ingredient1:ingredient6, ~sub(".*dashes", '', .))) %>%
  mutate(across(ingredient1:ingredient6, ~sub(".*dash", '', .))) %>%
  mutate(across(ingredient1:ingredient6, ~sub(".*drops", '', .))) %>%
  mutate(across(ingredient1:ingredient6, ~sub(".*drop", '', .))) %>%
  mutate(across(ingredient1:ingredient6, ~sub(".*bsp", '', .))) %>%
  mutate(across(ingredient1:ingredient6, ~str_replace(., " \\s*\\([^\\)]+\\)", "")))

View(cocktail_bar_cleaned)
```

Making it tidy

```{r}
cocktail_bar_cleaned <- cocktail_bar_cleaned %>%
  pivot_longer(cols=8:13, names_to = "ingredients", values_drop_na=TRUE)
```

Making them all lower case

```{r}
cocktail_bar_cleaned <- cocktail_bar_cleaned %>%
  mutate(value = tolower(value))
```

**3. Q: All ingredients in alphabetical order**
3. A: For this I am working in a list

Cleaining the white spaces

```{r}
ingredients_list<- trimws(cocktail_bar_cleaned$value, which = c("left"), whitespace = "[ \t\r\n]")
```

Making sure there are only one from each and arranging them

```{r}
ingredients_list_unique <- unique(ingredients_list)
ingredients_list_unique_arranged <- str_sort(ingredients_list_unique)
```

Here they are ordered:

```{r}
ingredients_list_unique_arranged
```

**4. Q: Number of unique ingredients**
4. A. There are 61.
```{r}
ingredients_list_unique_arranged_df <- as.data.frame(ingredients_list_unique_arranged)
nrow(ingredients_list_unique_arranged_df)
```

**5. Q: What are the top 10 ingredients?**

```{r}
top_ingredients <- cocktail_bar_cleaned %>%
mutate(value = fct_lump_n(value, n=10)) %>%
filter (value != "Other") %>%
  count(value) %>%
  arrange(n)
```

5. A: Here are the top 11 ingredients

```{r}
top_ingredients
```
**6. Q: Which cocktail(s) has/have the most ingredients?**

```{r}
cocktail_bar_cleaned$name <- as.factor(cocktail_bar_cleaned$name)
most_ingredients <- cocktail_bar_cleaned %>%
  mutate(name = fct_lump_n(name, n=1)) %>%
  filter (name != "Other") %>%
  count(name)
```

6. A: Here are the cocktails with the most (6) ingredients

```{r}
most_ingredients
```

**7. Q: How many ingredients appear in only one cocktail (rare ingredient)?**

```{r}
cocktail_bar_cleaned %>%
  count(value)

rare_ingredients <- cocktail_bar_cleaned %>%
  count(value)%>%
  filter (n == "1")
```

7. A: There are 29 ingredients which only appear in one cocktail.

```{r}
nrow(rare_ingredients)
```

**8. Q: Which cocktail has an ingredient that is only used in one cocktail?**

```{r}
cocktails_with_rares <- cocktail_bar_cleaned %>%
  mutate(value = fct_lump_n(value, n=-1)) %>%
  filter (value != "Other") %>%
  count(name)
```

8. A: Here they are:

```{r}
print(cocktails_with_rares$name)
```

**9. Q: What are the cocktails without rare ingredients?**

9. A: I have tried several things here, but I was not successful 

```{r}
#cocktails_with_rares_names <- as.factor(cocktails_with_rares$name) 

#rares <- as.data.frame(cocktails_with_rares$name)

#cocktails_without_rares <- cocktail_bar_cleaned %>%
  #pivot_wider(names_from = ingredients)

#all_cocktails <- as.factor(cocktails_without_rares$name)
#all_cocktails <- as.data.frame(all_cocktails)

#filter (all_cocktails != rares)
```

**10. Q. Create a cheat sheet for the bartender!**

```{r}
cocktail_bar_character <- as.character(cocktail_bar_cleaned$value)

cocktail_bar_cheatsheet <- cocktail_bar_cleaned %>%
  pivot_wider(names_from = value) 

cocktail_bar_cheatsheet[ , 10:71 ][ cocktail_bar_cheatsheet[ , 10:71 ] != "NA" ] <- "X"

cocktail_bar_cheatsheet[is.na(cocktail_bar_cheatsheet)] <- ""
```

10. A: Here is the cheatsheet

```{r}
cocktail_bar_cheatsheet
```
